How do changes in ex vary between countries?

### 2022-03-03 draft_paper.Rmd
Lu testing making edits in R stuio (line 14)
additional text please could you add more to line 467 on the difference between decades? I wonder what is easiest/most useful to include?

This notebook will compare annual changes in life expectancy at both birth ($e_0$) and at age 65 ($e_{65}$) from 1980 to 2019 (or the latest available year) in England and Wales, Scotland, France, Italy, Spains and Germany. Germany includes a 'Synthetic Germany' created from a weighted average of East and West German data for common years. The method for creating this 'Synthetic Germany' is covered in the methods section in brief and in Appendix X for more detail.

@@

- England & Wales (`GBRTENW`)
- Scotland (`GBR_SCO`)
- UK as a whole (`GBR_UK`)
- France (`FRANTP`)
- Spain (`ESP`)
- Italy (`ITA`)
- Germany
   - Total Germany (`DEUTNP`)
   - East Germany (`DEUTE`)
   - West Germany (`DEUTW`)
   - Simulated/Synthetic Germany

In order to explore the hypothesis that the recent stallin in life expectancy, most marked in England and Wales, was an expected 'correction' after a decade of unusually fast life expectancy improvements, we have selected the years from 1980 onwards.

@@

@@

## Descriptive Results /Starter Eyeballing

@@

The following figures show life expectancy at birth and at age 65 for selected nations from 1980 to the latest available year.

Here it is at birth...

@@

Here it is at age 65

@@

Scotland looks pretty bad compared with the other countries.

@@

Next, we compare the annual change in life expectancy at birth and at aged 65 years for the 6 nations from 1980 to the latest available year, by sex.

@@

- [ ] make smoother lines coloured by sex
- [ ] use consistent colours for all graphs showing separate sex results

```{r, echo = FALSE, warning=FALSE, message = FALSE, fig.height=8, fig.width=8}
change_in_ex_selected_countries %>% 
  filter(x == 0) %>% 
  left_join(country_code_lookup) %>% 
  mutate(country = factor(country, levels = c("England & Wales", "Scotland", "East Germany", "West Germany", "Germany", "Synthetic Germany", "Spain", "France", "Italy"))) %>% 
  filter(between(year, 1980, 2020)) %>% 
  ggplot(aes(x = year, y = delta_ex)) + 
  geom_point(alpha = 0.2) + 
  stat_smooth() + 
  facet_grid(country~sex) + 
  geom_hline(yintercept = 0) + 
  theme(strip.text.y = element_text(angle = 0)) + 
  labs(
    x = "Year",
    y = "Year-on-year change in life expectancy",
    title = "Annual change in life expectancy at birth, selected countries",
    subtitle = "Blue line: nonlinear smoother",
    caption =  "Source: Human Mortality Database. Synthetic Germany based on 20% East/80% West German population weighting"
  )

@@

Are the differences by decade statistically significant?

>>>>

Stalling of life expectancy improvements occurred in the UK from 2010 onwards. It was also seen in some other Western European countries. While the reason and severity of this is debated, some also contend the stalling of improvements seen is artefact following an unusual decade of improvement in 2000s (Murphy 2021).

-[ ] Introduction to be written once methods/results discussed and decided on

This notebook will compare annual changes in life expectancy at both birth ($e_0$) and at age 65 ($e_{65}$) from 1980 to 2019 (or the latest available year) in England and Wales, Scotland, France, Italy, Spains and Germany. Germany includes a 'Synthetic Germany' created from a weighted average of East and West German data for common years. The method for creating this 'Synthetic Germany' is covered in the methods section in brief and in Appendix X for more detail. In order to explore the hypothesis that the recent stalling in life expectancy, most marked in England and Wales, was an expected 'correction' after a decade of unusually fast life expectancy improvements, we have selected the years from 1980 onwards.

@@

- England & Wales (`GBRTENW`)
- Scotland (`GBR_SCO`)
- UK as a whole (`GBR_UK`) #we gaven't included UK but can do
- France (`FRANTP`)
- Spain (`ESP`)
- Italy (`ITA`)
- Germany
   - Total Germany (`DEUTNP`)
   - East Germany (`DEUTE`)
   - West Germany (`DEUTW`)
   - Simulated/Synthetic Germany

@@

We then compared whether the average between the decades was significant, using a regression model.

@@

## Descriptive Results

@@

The following figures show life expectancy at birth and at age 65 for selected nations from 1980 to the latest available year for males and females.

@@

@@

@@

Next, we compare the annual change in life expectancy at birth and at aged 65 years for the nations from 1980 to the latest available year, by sex.

@@

- [ ] make smoother lines coloured by sex
- [X] use consistent colours for all graphs showing separate sex results

```{r, echo = FALSE, warning=FALSE, message = FALSE, fig.height=8, fig.width=8}
change_in_ex_selected_countries %>% 
  filter(x == 0) %>% 
  left_join(country_code_lookup) %>% 
  mutate(country = factor(country, levels = c("England & Wales", "Scotland", "East Germany", "West Germany", "Germany", "Synthetic Germany", "Spain", "France", "Italy"))) %>% 
  filter(between(year, 1980, 2020)) %>% 
  ggplot(aes(x = year, y = delta_ex, colour=sex)) + 
  geom_point(alpha = 0.2) + 
  stat_smooth() + 
  facet_grid(country~sex) + 
  geom_hline(yintercept = 0) + 
  theme(strip.text.y = element_text(angle = 0)) + 
  labs(
    x = "Year",
    y = "Year-on-year change in life expectancy",
    title = "Annual change in life expectancy at birth, selected countries",
    subtitle = "Blue line: nonlinear smoother",
    caption =  "Source: Human Mortality Database. Synthetic Germany based on 20% East/80% West German population weighting"
  )

@@

####Are the differences by decade statistically significant? 
To assess whether the changes by decade are statistically significant, we carried out [Jon please could you write a little more here?]
We found the only significant change decade-on-decades was for East Germany for females at birth and at aged 65 years...
